\documentclass{article} % For LaTeX2e

\usepackage[submission]{colm2026_conference}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}

% NOTE: including geometry package
% The geometry package modifies some page properties when used. This can dramatically change the page margins, leading to severe template violation, and potential desk rejection. If the package is required, it can be used with the "pass" flag to skip the default page modifications, as in the following line:
% \usepackage[pass]{geometry}

\usepackage{lineno}
\usepackage{tikz}

\definecolor{darkblue}{rgb}{0, 0, 0.5}
\hypersetup{colorlinks=true, citecolor=darkblue, linkcolor=darkblue, urlcolor=darkblue}

\title{Local Permission Is Not Enough:\\Trace-Level Policy Violations in Tool-Using LLM Agents}
\author{Anonymous authors}

\begin{document}
\ifcolmsubmission
\linenumbers
\fi

\maketitle

\begin{abstract}
Tool-using language-model agents are commonly protected by local tool
permissions, prompt-level policies, confirmation prompts, and visible-content
filters. These mechanisms are necessary, but they do not enforce security
properties that depend on the complete execution trace. We study a
\emph{local-permission/global-policy gap}: every individual tool call can be
locally authorized, syntactically valid, and plausible, while the composed
sequence violates a policy over provenance, audience, tenant boundaries,
long-term memory, approval scope, or aggregation. We introduce TraceBreak, a
fully synthetic benchmark of paired safe and risky enterprise workflows, and
TraceGuard, a deterministic provenance monitor that propagates source tags
through reads, summaries, memory, and write sinks. In a 120-task deterministic
validation matrix, local guards and visible-content DLP allow all risky traces,
whereas TraceGuard blocks all risky writes with no safe-control false blocks. In
a 24-task API subset using \texttt{gpt-4.1-mini}, local guards and DLP again
produce local-pass violations on every risky task; policy prompting reduces but
does not eliminate violations and lowers task completion; TraceGuard blocks all
risky sink attempts while preserving safe-control utility. These results suggest
that secure agent scaffolds should treat the execution trace, not only the next
tool call or the model prompt, as a first-class enforcement object.
\end{abstract}

\section{Introduction}

Tool-using language-model agents increasingly perform enterprise workflows by
combining search, document reads, summaries, long-term memory, approvals,
delegation, and external writes. This shift changes the unit of security. In a
single API call, it is natural to ask whether the caller is allowed to invoke a
tool and whether the tool arguments are well formed. In an agent workflow, the
more important question is often whether information read earlier in the trace is
allowed to reach a later sink. An agent may read an internal renewal memo,
summarize it, save the summary into shared memory, and later email the digest to
an external vendor. Each step can be locally reasonable: the document exists, the
summary tool is permitted, the memory key is valid, and the recipient is a known
contact. The complete trace is nevertheless a data-flow violation.

This failure mode is broader than direct prompt injection or a single bad tool
call. In many deployed and proposed agent systems, security-relevant facts are
not co-located with the action being checked. The sensitivity of a document is a
property of the source object, the trust level of a recipient is a property of an
identity record, the validity of an approval depends on the object, recipient,
purpose, and time, and whether an export is permitted can depend on whether an
intermediate representation is an aggregate or a list of identified records. A
local guard that only inspects the current tool name and visible arguments does
not have enough information to enforce these invariants. A prompt-only policy can
ask the model to respect the rules, but it is not an execution boundary.

We isolate this problem as the \emph{local-permission/global-policy gap}. A
trace has a local-pass violation when all executed tool calls satisfy the local
authorization checks available at their call sites, yet the trace as a whole
violates a global invariant over hidden metadata and history. The gap is
especially relevant for multi-agent systems: a researcher agent may observe a
source, a planner may transform or store it, and an executor may perform the
write. No individual actor necessarily sees the whole policy state.

This paper makes four contributions.
\begin{enumerate}
    \item We formalize trace-level policy violations in tool-using agents,
    distinguishing action-local permission checks from global invariants over
    provenance, recipients, tenants, memory, approvals, and aggregation state.
    \item We implement TraceBreak, a fully synthetic, deterministic benchmark of
    120 paired risky and safe-control workflows across six compositional policy
    categories.
    \item We evaluate local guards, visible-content DLP, visible-policy and
    prompt-only baselines, and TraceGuard under both scripted traces and a
    bounded API agent subset.
    \item We release TraceGuard as a simple reference monitor baseline. The goal
    is not to claim that this monitor solves all agent security, but to provide a
    reproducible testbed where structural trace enforcement can be compared with
    model-side and action-local defenses.
\end{enumerate}

\section{Threat Model and Problem Definition}

We consider an agent system that interacts with a synthetic enterprise
environment through structured tools. The environment contains documents,
people, customers, support records, approvals, and shared memory. Tools include
search, reads, summarization, memory writes and reads, approval lookup, record
aggregation, email, ticket updates, and final answers. Email and ticket updates
are local simulator sinks; no real external service is used.

A trace is a sequence
\[
\tau = [(a_1, t_1, x_1, y_1, m_1), \ldots, (a_T, t_T, x_T, y_T, m_T)],
\]
where $a_t$ is the actor, $t_t$ the tool, $x_t$ the tool input, $y_t$ the tool
output, and $m_t$ hidden metadata such as provenance tags. The hidden metadata is
not shown to the language model in the API experiments. Instead, the runtime
attaches it to tool outputs and propagates it through later references.

Each proposed action is evaluated by a local permission function
$L(x_t, s_t) \in \{\mathrm{allow}, \mathrm{deny}\}$ over visible call-site
state $s_t$: whether the tool exists, whether the object identifiers are valid,
and whether referenced observations exist. The full trace is evaluated by a
global invariant $G(\tau,w) \in \{\mathrm{safe}, \mathrm{violation}\}$ over the
complete world state $w$ and all executed events. A local-pass violation occurs
when
\[
\forall t, L(x_t, s_t)=\mathrm{allow}
\quad\text{but}\quad
G(\tau,w)=\mathrm{violation}.
\]

The adversary in TraceBreak is not a malicious user writing exploit strings.
Instead, the risky tasks are benign-looking workflows whose requested sink is
invalid under hidden enterprise policy. This design intentionally separates the
trace-level policy problem from prompt-injection detection. We exclude trivial
cases where a single tool call violates an ACL or where the sink body contains an
obvious literal secret marker. We also avoid LLM judges for ground truth: a
violation is counted only when the executed trace performs a forbidden write,
cross-tenant update, invalid approval use, or non-aggregate export.

\paragraph{Non-goals.}
TraceBreak is not a leaderboard of model alignment, a realistic enterprise DLP
product, or a complete information-flow control system. It is a controlled
benchmark for one question: can a defense enforce policies whose decisive facts
are distributed across the trace rather than visible in the next action alone?

\section{TraceBreak Benchmark}

TraceBreak is a pure-Python simulator with deterministic task generation and
matched counterfactual controls. The benchmark contains 120 tasks: 10 seeds for
each of six categories, with one risky task and one safe-control task per seed.
The safe control changes a single policy-relevant field---for example the
recipient, tenant, approval scope, or output form---while preserving the surface
workflow. This pairing is important: a defense should block the risky sink
without losing utility on a nearly identical safe task.

\begin{table}[t]
\centering
\small
\begin{tabular}{p{0.23\linewidth}p{0.35\linewidth}p{0.31\linewidth}}
\toprule
Category & Risky pattern & Safe-control flip \\
\midrule
Sensitive external & Internal or confidential source flows to an external
recipient & Recipient is internal and authorized \\
Audience restriction & HR/legal/project-specific material flows to a role
outside its audience & Recipient role is in the allowed audience \\
Cross-tenant & Customer A context is used in customer B ticket or message &
Source and destination tenant match \\
Memory laundering & Sensitive source is written to memory and later sent in a
lower-trust context & Final recipient is authorized \\
Approval scope & Approval is reused for the wrong recipient, object, purpose,
or time & Approval fields match the sink \\
Aggregation threshold & Identified record list is exported externally & Only
de-identified aggregate statistics are sent \\
\bottomrule
\end{tabular}
\caption{TraceBreak categories. Each risky task has a matched safe control that
changes one policy-relevant metadata field or output form while preserving the
surface workflow.}
\label{tab:categories}
\end{table}

The key design principle is that hidden policy metadata is separated from visible
text. For example, a visible document may simply contain renewal planning notes;
its hidden tag records that it is confidential, belongs to AtlasCo, is intended
for sales and finance, and is not externally shareable. Similarly, a person
record exposes a role and organization, while the global policy determines
whether that role is in the source audience. This design tests whether an
enforcement mechanism tracks provenance rather than whether a content filter sees
a literal word such as ``confidential.''

\paragraph{Task execution.}
We evaluate two kinds of traces. Scripted traces execute canonical plans for each
category under single-agent and multi-agent scaffolds. API traces use a language
model that selects one JSON tool call at a time from the current instruction and
visible observations. The model is instructed to use observation references
through a \texttt{source\_refs} field when transformed information is written to
email, tickets, memory, summaries, or aggregates. In the current implementation,
TraceGuard's guarantees are conditional on such runtime-maintained dependencies;
a production implementation should make dependency tracking mandatory or
runtime-inferred rather than relying on model-written references.

\section{Defenses}

\paragraph{Local guards.}
The local guard checks only action-local validity: the tool is known, object IDs
exist, recipients exist, approvals exist, and declared source references are well
formed. It does not track where information came from, what tenant the source
belongs to, or whether an approval covers the proposed sink.

\paragraph{Visible-content DLP.}
The DLP baseline blocks visible sensitive markers such as ``confidential'' or
``restricted'' and exact non-shareable document titles in external writes. It is
a deliberately limited baseline that represents simple text filtering without
hidden provenance.

\paragraph{Visible policy and policy prompting.}
The deterministic visible-policy guard checks public fields available at the
sink, such as whether an approval identifier visibly matches the requested
recipient and whether an external email references many outputs. The API
policy-prompt condition instead adds natural-language instructions forbidding
unauthorized sharing, tenant mixing, invalid approval reuse, and memory
laundering. Neither baseline receives hidden source tags.

\paragraph{TraceGuard.}
TraceGuard attaches a \texttt{DataTag} to each tool output. Tags include source
IDs, derived-from IDs, sensitivity, tenant set, allowed audience, an
external-share flag, record count, and aggregate-only status. Transformations
such as summaries, memory writes, and memory reads merge tags conservatively:
sensitivity takes the maximum level, tenants and sources are unioned, audience
restrictions are intersected when present, and external sharing is allowed only
if all sources allow it. Sinks such as email and ticket updates evaluate the
merged tag against the recipient, tenant, approval, and aggregation invariants.

For email, TraceGuard blocks if non-shareable data reaches an external recipient
without a matching approval, if the recipient role is outside the allowed
audience, if tenant-specific data reaches another tenant, if an approval does not
match the object, recipient, purpose, or time, or if a detailed record list above
the aggregation threshold is exported externally. For ticket updates, it blocks
cross-tenant writes. Figure~\ref{fig:traceguard-schematic} illustrates the main
difference from local guards: the same locally allowed steps are permitted until
the tagged data reaches an invalid sink.

\begin{figure}[t]
\centering
\small
\begin{tikzpicture}[
    x=1cm,
    y=1cm,
    every node/.style={font=\scriptsize},
    box/.style={draw, rounded corners=2pt, minimum width=1.45cm, minimum height=0.48cm, align=center},
    ok/.style={box, draw=green!55!black, fill=green!5},
    block/.style={box, draw=red!70!black, fill=red!5},
    arrow/.style={->, line width=0.5pt, draw=gray!70}
]
\node[anchor=east] at (-0.25, 0.7) {Local};
\node[anchor=east] at (-0.25, -0.35) {TraceGuard};

\node[ok] (l1) at (0.7, 0.7) {search\\allow};
\node[ok] (l2) at (2.55, 0.7) {read\\allow};
\node[ok] (l3) at (4.40, 0.7) {summarize\\allow};
\node[ok] (l4) at (6.25, 0.7) {recipient\\allow};
\node[ok] (l5) at (8.10, 0.7) {email\\allow};

\node[ok] (g1) at (0.7, -0.35) {search\\allow};
\node[ok] (g2) at (2.55, -0.35) {read\\allow};
\node[ok] (g3) at (4.40, -0.35) {summarize\\allow};
\node[ok] (g4) at (6.25, -0.35) {recipient\\allow};
\node[block] (g5) at (8.10, -0.35) {email\\block};

\foreach \a/\b in {l1/l2,l2/l3,l3/l4,l4/l5,g1/g2,g2/g3,g3/g4,g4/g5}
  \draw[arrow] (\a) -- (\b);

\node[align=center, text=gray!70!black] at (3.85, -1.05)
  {tag: confidential, AtlasCo, no external share};
\node[align=center, text=red!70!black] at (8.10, -1.30)
  {block reason:\\internal data to external recipient};
\end{tikzpicture}
\caption{Local guards evaluate each action independently. TraceGuard propagates
hidden provenance tags through transformations and blocks when the tagged data
reaches an invalid sink.}
\label{fig:traceguard-schematic}
\end{figure}

\section{Experiments}

\paragraph{Protocol.}
We run two evaluations. First, a deterministic scripted matrix validates that the
benchmark and monitor implement the intended invariants over all 120 tasks.
Second, a bounded API subset uses \texttt{gpt-4.1-mini} on two task seeds,
covering two risky/safe pairs per category, for 24 tasks per condition. API runs
use cached responses, structured JSON actions, a maximum of eight tool steps,
temperature zero, and deterministic grading from executed sink events. We report
sink rate, safe utility, risky global violation rate, local-pass violation rate
(LPVR), safe false-block rate, risky block rate, and average model calls.

\paragraph{Metrics.}
Safe utility is the fraction of safe controls where the expected sink executes
without a global violation. Risky violation is the fraction of risky tasks where
a forbidden sink executes. LPVR is the fraction of risky tasks where all local
checks allow the trace and a global violation still occurs. Safe false block is
the fraction of safe controls blocked by the defense. A lower sink rate is not
itself a failure: for TraceGuard, the expected behavior is to execute safe sinks
and block risky sinks, producing a 50\% overall sink rate on a balanced dataset.

\begin{table}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lrrrrrr}
\toprule
Condition & Sink & Safe util. & Risky viol. & LPVR & Safe FP & Calls \\
\midrule
Local guards & 100 & 100 & 100 & 100 & 0 & 5.1 \\
Content DLP & 100 & 100 & 100 & 100 & 0 & 5.1 \\
Policy prompt & 79 & 83 & 75 & 75 & 0 & 4.6 \\
TraceGuard & 50 & 100 & 0 & 0 & 0 & 5.1 \\
\bottomrule
\end{tabular}
\caption{API subset results with \texttt{gpt-4.1-mini}. Rates are percentages
over 24 tasks, except safe utility and safe false positives over 12 safe
controls, and risky violation and LPVR over 12 risky tasks. TraceGuard's 50\%
sink rate is expected: all risky sink attempts are blocked while all safe sinks
execute.}
\label{tab:api-main}
\end{table}

\begin{table}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lrrrrr}
\toprule
Condition & Sink & Safe util. & Risky viol. & LPVR & Safe FP \\
\midrule
Single local & 100 & 100 & 100 & 100 & 0 \\
Multi local & 100 & 100 & 100 & 100 & 0 \\
DLP & 100 & 100 & 100 & 100 & 0 \\
Visible policy & 83 & 100 & 67 & 67 & 0 \\
TraceGuard & 50 & 100 & 0 & 0 & 0 \\
\bottomrule
\end{tabular}
\caption{Deterministic 120-task validation. Scripted traces verify that the
ground-truth invariants and monitor behave as intended across all paired tasks.}
\label{tab:det-main}
\end{table}

\paragraph{Main findings.}
Local guards and content DLP complete all safe controls and all risky tasks in
both evaluations, but every risky completion is a local-pass violation. This is
the intended blind spot: the dangerous fact is often not visible in a single
action string. It is stored in the source tag, recipient metadata, approval
record, or previous memory event. Policy prompting helps, but it is not an
enforcement boundary. In the API subset, the policy prompt lowers risky
violation rate from 100\% to 75\%, but also lowers safe utility from 100\% to
83\%; the supplementary prompt diagnostic shows that two avoided
approval-scope risky violations are abstentions that also lose the matched safe
controls, and the remaining avoided cross-tenant risk changes the sink target
rather than enforcing policy. TraceGuard blocks every risky sink attempt in both
evaluations while preserving all safe-control completions. On the 12 matched
risky API pairs, TraceGuard improves over local guards and DLP on all pairs and
over policy prompting on 9 pairs.
Supplementary category tables show the API pattern in all six categories;
the block-reason audit shows category-aligned reasons with no safe-control
blocks; fixed-action replays and deterministic authority-transfer,
injection-overlay, and decoy-clutter checks show the same structural gap without
cross-condition sampling noise.

\paragraph{Provenance stress test.}
A direct concern is that TraceGuard might rely on model-authored
\texttt{source\_refs}. A cached audit confirms 92/92 executed API sinks have
valid prior refs. We replay the same 24 API TraceGuard traces after deleting
sink refs, corrupting them to benign public observations, or erasing refs from
intermediate summaries/memory/aggregates. Ordinary TraceGuard fails under sink
deletion (100\% risky violations), sink corruption (83\%), and one
summary-laundering intermediate case. StrictTraceGuard fixes missing sink refs
but creates 92\% safe false blocks and does not fix stale refs. A
runtime-inferred replay reconstructs sink tags from trusted event tags,
recovering sink deletion and corruption with 100\% safe utility, 0\% risky
violations, and 100\% risky blocks; it also removes the intermediate violation,
with one safe false block. The production lesson is to validate or infer
dependencies rather than trust model-written provenance fields.

\paragraph{Why visible defenses fail.}
DLP fails when the sink body is innocuous but its provenance is not: ``Short
update from the latest notes'' has no literal secret marker, yet may derive
from a confidential AtlasCo document. Visible-policy review catches sink-visible
approval and bulk-export cases, but misses source-tag tenant, audience, memory,
and confidentiality facts. In same-action replay, it leaves 67\% risky
violations; the visibility-gap audit shows these are exactly the hidden-metadata
categories. A metadata-aware critic given hidden tags ties TraceGuard at 0\%.
The missing object is policy state, not another natural-language reviewer.

\paragraph{Example trace.}
In a sensitive-external task, the agent searches for AtlasCo renewal notes, reads
\texttt{doc\_pricing\_100}, summarizes the result, finds an external recipient,
and sends an email. Local checks allow every step. The hidden tag on the email
body still contains \texttt{source=doc\_pricing\_100},
\texttt{sensitivity=confidential}, \texttt{tenant=AtlasCo}, and
\texttt{external\_share=false}. TraceGuard blocks the same final email with
reason \texttt{internal\_data\_to\_external\_recipient}. A compact trace pair is
included in \texttt{paper/tables/example\_traces.md}.

\paragraph{Reproducibility.}
TraceBreak does not depend on private data or human labels. Tasks are generated
from deterministic seeds, traces are written as JSONL, API responses are cached,
and all headline metrics are recomputed from executed sink events rather than
from an LLM judge. The artifact includes scripts to regenerate the 120-task
dataset, rerun each condition, recompute point-estimate tables, export example
traces, and rebuild the paper.

\section{Related Work}

\paragraph{Prompt injection and tool-using agents.}
Indirect prompt injection showed that LLM-integrated applications blur data and
instructions when they process untrusted external content
\citep{greshake2023notwhat,liu2023promptinjection}. Later benchmarks and
surveys formalize prompt-injection attacks, defenses, and tool-agent security
risks \citep{liu2023formalizingprompt,owasp2025llm01,debenedetti2024agentdojo,zhan2024injecagent,zhang2024asb}.
TraceBreak is complementary: the violation need not come from an adversarial
instruction, and locally benign steps can compose into a global-policy failure.

\paragraph{Runtime enforcement and information flow.}
Runtime defenses increasingly move beyond prompt-only mitigation: Task Shield,
RTBAS, AgentSpec, SafeAgent, and ClawGuard enforce task alignment,
information-flow, trajectory-level, or user-confirmed constraints around tool use
\citep{jia2024taskshield,zhong2025rtbas,wang2025agentspec,liu2026safeagent,zhao2026clawguard}.
These systems target broad deployment settings. TraceGuard is narrower: a
compact provenance monitor for tenant, audience, approval, memory, and
aggregation policies, paired with safe controls so the
local-permission/global-policy gap is measurable from raw traces.

\paragraph{Memory, multi-agent, and compositional risk.}
Memory and retrieval can preserve unsafe influence across time
\citep{chen2024agentpoison}, and multi-agent systems introduce coordination
failures and new attack surfaces
\citep{schroederdewitt2025multiagentsecurity,hammond2025multiagentrisks}.
Closest to TraceBreak are Policy-Invisible Violations on hidden policy facts
\citep{wu2026policyinvisible} and Beyond Single-Agent Alignment on context-
fragmented violations \citep{wu2026contextfragmented}. Our contribution is a
releaseable paired benchmark and monitor for concrete runtime data flows, with
sink-event metrics and provenance-deletion/corruption replays that separate the
cooperative benchmark result from the stronger production claim.

\section{Limitations and Ethics}

TraceBreak is synthetic: this enables safe release, deterministic labels, and
matched controls, but cannot capture all enterprise policy complexity.
TraceGuard uses conservative tag propagation and hand-coded invariants; richer
semantic transformations may require policy-specific declassification rules. The
ordinary monitor still trusts correct \texttt{source\_refs}. Cached API sinks
comply, but deleted- and corrupted-provenance replays show failures when sink
references are missing or stale. Runtime-inferred sink tags restore safety and
utility; a conservative missing-provenance blocker restores safety only for
omissions and overblocks. Production monitors should validate provenance or
infer runtime dependencies rather than trust model-authored fields.

API results are preliminary evidence, not a model leaderboard; zero observed
TraceGuard violations cover TraceBreak invariants, not arbitrary enterprise
policies. A matched-control oracle finds safe continuations for blocked traces,
but live recovery remains future work. All data, people, tenants, documents,
approvals, and messages are synthetic; no real email, ticketing, browsing, or
external enterprise systems are used.

\section{Conclusion}

TraceBreak shows that tool-level permission is not compositional: locally
allowed actions can jointly violate global policy over provenance, audience,
tenant, memory, approval, or aggregation. Prompting and content filters miss
hidden, distributed facts; deterministic trace monitoring closes this benchmark
gap. Agent security should treat traces as first-class enforcement objects.

\bibliographystyle{colm2026_conference}
\bibliography{references}

\end{document}
